% !TEX root = 02.tex
\include{common.tex}

\title{Formal verification - Tutorial 02}

\begin{document}

\maketitle

\section*{Linear temporal logic}

Fix a set of propositional variables $P$.
%
Recall that linear temporal logic (LTL) formulas over $P$
are defined by the following (minimalistic) grammar:
%
\begin{align*}
	\varphi, \psi ::= \true \sep \false \sep p \sep \neg \varphi \sep \varphi \wedge \varphi \sep \X \varphi \sep \varphi \U \psi,
\end{align*}
%
where $p \in P$.
%
We define the derived operators
%
\begin{align*}
	\F \varphi &= \true \U \varphi \\
	\G \varphi &= \neg \F \neg \varphi.
\end{align*}

\begin{exercise}
	Write LTL formulas for the following properties:
	\begin{enumerate}
		\item $p$ holds infinitely often;
		\item whenever $p$ holds, then $q$ holds afterwards;
		\item $p$ holds \emph{precisely} at every even step;
	\end{enumerate}
\end{exercise}

\begin{solution}
	To express that $p$ holds infinitely often,
	we can write $\G \F p$.
	The second property is expressed by $\G (p \to \F q)$.
	The last property can be expressed in a variety of ways.
	For instance, we can write
	%
	\begin{align*}
		p \wedge \G ((p \implies \X \neg p) \wedge (\neg p \implies \X p)). \qedhere
	\end{align*}
\end{solution}

\begin{exercise}
	Consider the property
	``for every prefix of the computation,
	the number of times $p$ holds is at most the number of times $q$ holds''.
	Is this property expressible in LTL?
\end{exercise}

\begin{solution}
	No, the stated property is not expressible in LTL.
	The reason is that LTL formulas can only express $\omega$-regular properties,
	and the property in question is not $\omega$-regular.
\end{solution}

\begin{exercise}
	Consider the property
	``$p$ holds at all even time steps'' (but possibly also at odd steps).
	Is this property expressible in LTL?
\end{exercise}

\begin{solution}
	This property is not expressible in LTL,
	but in order to show it, we need to use a more advanced tool (next tutorial).
\end{solution}

\section*{From LTL to alternating Büchi automata}

We recall the translation from LTL to alternating Büchi automata.
%
To this end, we first convert an LTL formula $\varphi$ to negation normal form,
i.e., we push negations inside until they only occur in front of propositional variables.
%
To achieve this, we introduce the De Morgan dual of the until operator, which is called the \emph{release} operator:
%
\begin{align*}
	\varphi \Release \psi &= \neg (\neg \varphi \U \neg \psi).
\end{align*}
%
Note that $\G \varphi$ is equivalent to $\false \Release \varphi$.

We will use the following fixpoint expansions of the until and release operators:
%
\begin{align*}
	\varphi \U \psi &= \psi \vee (\varphi \wedge \X (\varphi \U \psi)), \\
	\varphi \Release \psi &= \psi \wedge (\varphi \vee \X (\varphi \Release \psi)).
\end{align*}

Given an LTL formula $\varphi$ in negation normal form,
we construct an alternating Büchi automaton
$A = \tuple{\Sigma, Q, q_0, F, \delta}$ as follows.
%
States of the automaton $Q$ are the subformulas of $\varphi$,
to which we add $\true$ and $\false$.
%
The initial state is $q_0 = \set \varphi$
and the set of accepting states $F$ contains all the release formulas and $\true$.
%
The transition function is defined by induction
on the structure of states $\psi \in Q$.
%
The cases for propositional variables and positive Boolean connectives are straightforward:
%
\begin{align*}
	\delta(\true, a) &= \true, \\
	\delta(\false, a) &= \false, \\
	\delta(p, a) &= \true \text{ if } p \in a, \text{ and } \false \text{ otherwise}, \\
	\delta(\neg p, a) &= \true \text{ if } p \not\in a, \text{ and } \false \text{ otherwise}, \\
	\delta(\psi \wedge \chi, a) &= \delta(\psi, a) \wedge \delta(\chi, a), \\
	\delta(\psi \vee \chi, a) &= \delta(\psi, a) \vee \delta(\chi, a).
\end{align*}
%
The most interesting cases are those for the temporal operators.
%
\begin{align*}
	\delta(\X \psi, a)
		&= \psi, \\
	\delta(\psi \U \chi, a)
		&= \delta(\chi, a) \vee (\delta(\psi, a) \wedge \psi \U \chi), \\
	\delta(\psi \Release \chi, a)
		&= \delta(\chi, a) \wedge (\delta(\psi, a) \vee \psi \Release \chi).
\end{align*}
%
For until and release we have applied the transition function $\delta(\cdot, a)$
to the fixpoint expansions given above.

\begin{exercise}
	Apply the translation above
	to construct an alternating Büchi automaton
	for the LTL formula $\G \F p$.
\end{exercise}

\begin{exercise}
	Using the automaton above, how can we construct an automaton for $\G \F p \wedge \G \F q$?
\end{exercise}

\end{document}
