% !TEX root = 04.tex
\include{common.tex}

\title{Formal verification - Tutorial 04}
\date{16/03/2026}

\begin{document}

\maketitle

\newcommand{\lossyto}{\Rightarrow}
\section*{Communicating finite-state machines}

Let $S$ be CFSM with (perfect) transition relation $\to$.

\begin{exercise}
	Show that the reachability problem for CFSMs is undecidable.
\end{exercise}

\begin{solution}
	We can simulate the tape of a Turing machine using a single channel.
	A special symbol is used to mark the position of the head.
	The head movement are simulated by rotations of the channel contents.
\end{solution}

We consider a model of CFSM comprising a finite set of processes $P$ (each with a finite set of local states)
and a finite set of channels $C$.
%
Each channel is a FIFO queue of messages connecting two processes (point-to-point communication).
%
The \emph{communication topology} of a CFSM is an undirected multigraph
where vertices are processes and (undirected) edges are channels.

\begin{exercise}
	Show that the reachability problem for CFSMs is undecidable when
    %
    \begin{enumerate}
        \item there are two processes $P$ and $Q$ and two channels $C = \set {c, d}$,
        where $c$ connects $P$ to $Q$ and $d$ connects $Q$ to $P$ (i.e., the communication topology is a cycle of length two);
        \item there are two processes $P$ and $Q$ and two ``parallel'' channels $C = \set {c, d}$
        both connecting $P$ to $Q$ (i.e., the communication topology is a multigraph with two parallel edges).
    \end{enumerate}
\end{exercise}
\begin{solution}
    1.~We show how to simulate a single channel $e$ (where the same process can send and receive)
    by the two channels $c, d$.
    Transmissions on $e$ are simulated by transmissions on $c$.
    Receptions on $e$ are simulated by receptions on $d$.
    Both these are taken care by process $P$.
    Process $Q$, which is fixed, just forwards messages from $c$ to $d$.

    2.~We reduce from the \emph{Post correspondence problem} (PCP), which is an undecidable problem.
    %
    An instance of this problem is given by a finite sequence of pairs of words
    $\tuple{u_1, v_1}, \dots, \tuple{u_n, v_n}$ over some alphabet $\Sigma$
    and asks whether there is a sequence of indices $i_1, \dots, i_k$ such that
    $u_{i_1} \cdots u_{i_k} = v_{i_1} \cdots v_{i_k}$.
    %
    Process $P$ guesses a sequence of indices $i_1, \dots, i_k$
    and sends $u_{i_1}, \dots, u_{i_k}$ on channel $c$
    and $v_{i_1}, \dots, v_{i_k}$ on channel $d$.
    %
    Process $Q$ receives messages from both channels and checks that they are the same.
\end{solution}

\begin{exercise}
	Show that the reachability problem for CFSMs is decidable \emph{over tree communication topologies}.
\end{exercise}

\begin{solution}
	When the communication topology is a tree,
	bounded channels suffice to reach all reachable configurations.
	%
	The idea is that we can always keep the channels to size zero or one,
	by always executing a transmision immediately followed by the corresponding reception.

    Another idea, is as follows.
    We assume w.l.o.g~that the channel alphabets are disjoint.
    For the leaf processes, we can compute the regular language of messages they can send or receive (but not both).
    %
    Inductively, consider an internal process $p$ connected to child processes $p_1, \dots, p_k$.
    For each we have inductively computed the regular languages $L_1 \subseteq M_1^*, \dots, L_k \subseteq M_k^*$ of channel actions.
    We compute the regular language $L \subseteq M$ of channel actions of $p$ to its parent process.
    %
    Let $L'$ be the shuffle of languages $L_1, \dots, L_k$ and $M^*$.
    Then $L$ is the projection to $M^*$
    of the language of all channel actions of $p$ intersected with $L'$.
\end{solution}

\section*{Well-quasi orders}

A binary relation $\preceq$ on a set $X$ 
is \textbf{quasi-order}
if it is reflexive and transitive.
An antichain is a subset of $X$ 
whose elements are pairwise incomparable w.r.t.~$\preceq$.

\begin{exercise}
    Let $(X, \preceq)$ be a quasi-order.
    Prove that the following conditions are equivalent.
    \begin{enumerate}%[label=($\alph*$)]
        \item[(a)] Any infinite sequence $a_1, a_2, \dots$ of elements of $X$
        contains a \emph{domination} $a_i \preceq a_j$
        for some indexes $i < j$.

        \item[(b)] $X$ does not contain any infinite antichain and is \emph{well-founded}
        (i.e., there are no infinite strictly decreasing sequences).

        \item[(c)] Any infinite sequence $a_1, a_2, \dots$ of elements of $X$
        contains an infinite nondecreasing subsequence, i.e., there are indices
        $i_1 < i_2 < \ldots$ with $a_{i_1} \preceq a_{i_2} \preceq \ldots$.

        \item[(d)] Every upward closed set is the upward closure of a finite set.
        (The \emph{upward closure} of a set $X$
        is the set of elements that dominate some element of $X$;
        $X$ is \emph{upward closed} if it is equal to its upward closure.)

        \item[(e)] Every nondecreasing chain of upward closed sets $U_1 \subseteq U_2 \subseteq \cdots \subseteq X$ is finite.
    \end{enumerate}
\end{exercise}
\noindent
Whenever any of the conditions above holds we call 
the relation $\preceq$ a \textbf{well quasi-order} or \textbf{WQO}.
%
\begin{solution}
    We show the implication from (b) to (c) using the infinite Ramsey theorem.
    Consider an infinite sequence $a_1, a_2, \dots$ of elements of $X$.
    We colour the edges of the complete undirected graph on $\set{a_1, a_2, \dots}$
    with three colours $\set{A, B, C}$: we colour the edge $\set{a_i, a_j}$ with $i < j$
    with $A$ if $a_i \preceq a_j$,
    with $B$ if $a_i \succ a_j$,
    and with $C$ if $a_i$ and $a_j$ are incomparable.
    %
    By the infinite Ramsey theorem, there is an infinite monochromatic clique,
    that is a subsequence $a_{i_1}, a_{i_2}, \dots$ such that all edges $\set{a_{i_j}, a_{i_k}}$ with $j < k$ have the same colour.
    By assumption (b), this color cannot be $B$,
    since otherwise we would have an infinite strictly decreasing sequence.
    It cannot be $C$ either, since otherwise we would have an infinite antichain.
    Hence, the color is $A$ and we have an infinite nondecreasing subsequence, as required.
\end{solution}

\begin{exercise}
	Show that every downward-closed language of words (w.r.t.~the subword ordering) is regular.
\end{exercise}

% \begin{solution}
% \end{solution}

\section*{Lossy channel systems}

Let $S$ be a CFSM with (perfect) transition relation $\to$
and lossy transition relation $\lossyto$.
%
Recall that $c \lossyto d$ if there are configurations $c', d' \in \Conf$ such that
$c' \sqsubseteq c$, $c' \to d'$, and $d \sqsubseteq d'$.
%
Consider the following predecessor operators:
%
\begin{align*}
	\PreRel \to T
		&:= \setof {c \in \Conf}{\exists c' \in T, c \to c'}, \\
	\Pre T
		&:= \setof {c \in \Conf}{\exists c' \in T, c \Rightarrow c'},
\end{align*}
%
Notice that $\Pre T$ can be infinite even when $T$ only contains a single configuration.
%
\begin{exercise}
	Let $T$ be a regular set of configurations.
	Show that the following sets are effectively regular:
	\begin{itemize}
		\item $\PreRel \to T$,
		\item $\upwardclosure T$,
		\item $\Pre T$,
		\item $\PreStar T$.
	\end{itemize}
\end{exercise}

\begin{solution}
	The set $\PreRel \to T$ can be obtained from $T$ by regularity-preserving operations
	such as finite unions, concatenations, and right language quotients.
	%
	From a finite automaton recognising $T$,
	we can construct a finite automaton recognising $\upwardclosure T$ by adding
	self-loops on all states for all symbols.
	%
	Straight from the definitions, we have
	\begin{align*}
		\Pre T = \upwardclosure{\PreRel \to {\upwardclosure T}}. \qedhere
	\end{align*}
\end{solution}

\bibliographystyle{plainurl}
\bibliography{bibliography}

\end{document}